Static Analysis by Path Focusing

### 2011-05-13 report/report.tex
report + added counters in the code

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification aims at statically discover properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
	\\ \\  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation, Invariant
Generation}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

	\cite{GulwaniZ10} proposes a technique to compute transitive closure of a
	transition system, i.e compute the invariants for a loop having its semantic
	expressed by the transition system.

	A transition system for a control point $p$ is a DNF formula where each
	disjunct is the semantic of a path from $p$ to itself. It is of the form
	$$\bigvee_{1 \leq i \leq n} \tau_{p,i}$$ where $n$ is the number of
	paths, and $\tau_{p,i}$ is the semantic of the $i$-th path from $p$ to itself.

@@

The new algorithm we propose is described in Algorithm \ref{gulwani2}.
A difference is that here, we don't know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
compute this number of paths at the beginning of the algorithm.

@@

@@

	\emph{Path focusing} technique requires to decide the satisfiability of some
	SMT formula, and to extract a model when the formula is satisfiable, i.e an
	assignement of the variables so that the formula is \emph{true}. For doing
	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
	and Yices \cite{DutertreM06}. One can easily switch from one to another with
	an argument we give as parameter to the analyzer. In this way, we can
	compare the effectiveness of both SMT-solvers.

@@

	In all experiments, the set $P_R$ is chosen equal to $P_W \cup P_{ret} \cup
	P_{init}$,
	where $P_W$ is the set of widening points. $P_{init}$ is the set of starting
	points of the program, and $P_{ret}$ is the set of points that have no
	successors.

@@

@@

@@

	In this report, we proposed a new approach to guide the iterations of the
	analyzer, in order to distinguish the different paths and to treat them
	independently. The loss of precision due to widening and control-flow merges
	is then reduced. This technique allows us, for instance, to do narrowing
	steps right after widening, or to use acceleration techniques when a path
	actually is a simple loop.

	We implemented this technique into a small analyzer, 
	and it appears that in lots of cases, 
	the precision of the result is improved 
	compared to the classical abstract interpretation technique.
	We compared it with \emph{Lookahead Widening}, and our technique seems to
	perform better in the case of loops containing control-flow merges.

>>>>

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification aims at statically discover properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

	\cite{GulwaniZ10} proposes a technique to compute transitive closure of a
	loop, i.e compute the invariants for a loop having its semantic
	expressed by a transition system. The definition of a transition system is
	the following:

	\begin{definition}
	A transition system for a control point $p$ is a DNF formula (i.e a
	disjunction of conjunctions), where each
	disjunct is the semantic of a path from $p$ to itself. It is of the form
	$$\bigvee_{1 \leq i \leq n} \tau_{p,i}$$ where $n$ is the number of
	paths, and $\tau_{p,i}$ is the semantic of the $i$-th path from $p$ to
	itself.
	\end{definition}

@@

The new algorithm we propose is described in Algorithm \ref{gulwani2}.
A difference is that here, we don't know the number of paths. Then, the mapping
function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
easily compute this number of paths when computing the $\rho$ formula.

@@

\subsection{Diseq comparisons}

The set of elements $x$ satisfying $x \neq N$ is not convex: it is the union of
two intervals: $]-\infty, N[\  \cup\  ]N,+\infty[$. Since we use convex polyhedra to
represent the set of possible states of our program, we loose information when
the analysis comes across a \emph{diseq} operation.
One classical approach would be to compute a disjunctive invariant, by attaching
two different polyhedron to the BasicBlock. However, we prefered the following
technique: we transform all the operations of the form $x \neq y$ into
$x < y \vee x > y$, meaning that we separate the cases \emph{lesser than} and
\emph{greater than} into two distinct paths.

\begin{minipage}[c]{.29\linewidth}
\begin{C}
if (x != y) {
	...
}
\end{C}
\end{minipage} $\Rightarrow$
\begin{minipage}[c]{.69\linewidth}
\centering
\begin{C}
	if (x > y) {
	} else if (x < y) {
	} else {
		...
	}
\end{C}
\end{minipage}

To do so, one could implement an LLVM pass that transform the internal
representation according to this principle. However, this pass has not been
implemented in our analyzer, and this transformation has to be done manually.

@@

	\emph{Path focusing} technique requires to decide the satisfiability of some
	SMT formula, and to extract a model when the formula is satisfiable, i.e an
	assignement of the variables so that the formula is \emph{true}. For doing
	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
	and Yices \cite{DutertreM06}. One can easily switch from one to another with
	an argument we give as parameter to the analyzer. In this way, we can
	compare the efficiency of both SMT-solvers.

	\section{Limitations}

	Our analyzer could be extended in many ways. Indeed, it is intra-procedural,
	meaning that it does not consider that a variable may change after a
	function call. In addition, the return value of a function call is
	considered unconstrained. This limitation can be addressed by simply
	inlining the function call, provided the function is not recursive.
	There have been research to find efficient interprocedural analysis
	techniques \cite{Bourdoncle90,Bou92}.

	Additionnaly, our analyzer has no memory model:
\begin{itemize}
\item \emph{Store} instructions, that store a value in the memory, has no effect
in the analysis.
\item \emph{Load} instructions, that get a value $v$ in the memory and assign it
to a variable $x$, doesn't provide properties about the value $v$. Then, our
analyzer considers $x$ can take all values of its type.
\end{itemize}
In the case of local variables, the transformation of the internal
representation into SSA form removes the \emph{Load/Store} instructions.
The use of global variables instead of local ones will induce huge
loss of precision, since there will still be \emph{Load/Store} instructions in
the IA.

@@

	In all experiments, the set $P_R$ is chosen equals to $P_W \cup P_{ret} \cup
	P_{init}$,
	where $P_W$ is the set of widening points. $P_{init}$ is the set of starting
	points of the program, and $P_{ret}$ is the set of points that have no
	successors.

@@

ex6.c & 0.104s & $\infty$ & \begin{tabular}{l}
-y.0 + 51 >= 0 \\
y.0 >= 0 \\
x.0 - 51y.0 + 2550 >= 0 \\
x.0 >= 0
\end{tabular} & 0.032 & \begin{tabular}{l}
y.0 >= 0
\end{tabular} \\ \hline

ex8c.c & 191m50.935s & $\infty$ & \begin{tabular}{l}
c.0 + a.0 = 0 \\
b.0 + d.0 = 0\\
-b.0 - c.0 + 1 >= 0\\
-b.0 + c.0 + 1 >= 0\\
b.0 - c.0 + 1 >= 0\\
b.0 + c.0 + 1 >= 0\\
\end{tabular} & 0.928 & \begin{tabular}{l}
-b.0 + 1 >= 0\\
-b.0 + a.0 + 1 >= 0\\
-c.0 + 1 >= 0\\
-c.0 + d.0 + 1 >= 0\\
-a.0 + 1 >= 0\\
-d.0 + 1 >= 0\\
a.0 + 1 >= 0\\
c.0 - d.0 + 1 >= 0\\
c.0 + d.0 + 1 >= 0\\
b.0 - a.0 + 1 >= 0\\
b.0 + 1 >= 0
\end{tabular} \\ \hline

@@

To be completed\dots

\begin{table}[!h]
\begin{tabular}{|l||c|c|c|c|c|c|c|} \hline
Benchmark  & Gopan.c & Beyer.c & Boustr.c & ex8.cpp & ex9.cpp & test1.c & ex6.c\\ \hline \hline
paths (total) & 5 & 4 & 10 & 75626 & 8 & 3 & 5 \\ \hline
paths (computed) & 5 & 4 & 4 & 4 & 4 & 3 & 4\\ \hline
iterations  & 6 & 5 & 5 & 4 & 4 & 3 & 4\\ \hline
\end{tabular}
\caption{Number of path in the expanded multigraph, number of path the
algorithm actually computes, and total number of iterations}
\end{table}

To be completed\dots

@@

	Validation of softwares has become an essential domain in industry,
	especially those dealing with embedded systems, where a bug in the program
	can endanger the safety of people, or induce huge costs to fix. These
	domains constitute a privileged application field for formal verification
	techniques.
	Abstract Interpretation is a common used technique to do static analysis of
	programs. It aims to compute iteratively an over-approximation of the set of
	possible states of the program, and allows to prove properties such as
	``In this program, $x$ is always between 1 and 10'', ``The program never
	divides by zero'', etc.

	In this report, we proposed a new approach to guide the iterations of the
	analyzer, in order to distinguish the different paths of the program
	and to treat them
	independently. The loss of precision due to widening and control-flow merges
	is then reduced. This technique allows us, for instance, to do narrowing
	steps right after widening, or to use acceleration techniques when a path
	actually is a simple loop.

	Since distinguishing all the paths leads to an exponential blowup in time
	and memory, we do compute a path only when needed, otherwise the path is
	kept implicit. In this way, memory consumption is preserved, and the
	computation time is reduced, but still potentially exponential.
	In order to find which path we should compute, we use SMT-solving to find
	paths that makes the abstract interpretation analysis progress.

	We implemented this technique into a small analyzer dealing with C and C++
	programs, and run it on several
	benchmarks.
	It appears that in lots of cases, 
	the precision of the result is improved 
	compared to the classical abstract interpretation technique.
	We compared it with \emph{Lookahead Widening}, and our technique seems to
	perform better in the case of loops containing control-flow merges.

	Our technique performs better because of the distinction between
	paths. Then, encoding some properties, such as arithmetic overflows or
	aliases, into distinct paths in the control flow graph is one direction to
	explore.

	Our analyzer is able to deal with small- and middle-sized programs, but lose
	precision in the case of programs with pointers or several functions.
	It could be improved in many ways, in order to support
	for instance interprocedural analysis. A memory model could be added to
	handle programs with pointers and \emph{load/store} instructions.
	One could also try to compute disjunctive invariants using the technique
	described in this report.
	To be completed\dots
